Comparison of several sequential and parallel derivative-free global optimization algorithms

### 2018-11-29 2019/PAVT/pavt_paper.tex
PAVT: update

\begin{abstract}
This work considers several stochastic and deterministic derivative-free global optimization
algorithms. In the first part of the paper popular sequential open-source solvers are compared against
the Globalizer solver, which is developed at the Lobachevsky State University. The Globalizer is
designed to solve problems with black-box objectives satisfying the Lipschitz condition and shows
competitive performance with other similar solvers. The comparison is done on several sets of
challenging multi-extremal benchmark functions. The second part of this work is devoted to
comparison between the Globalizer and MIDACO solvers on systems with shared and distributed
memory. MIDACO is a state-of-the-art global solver included to the TOMLAB optimization
environment for MATLAB. Results of the benchmark show advantages of the Globalizer on small-
dimensional, but sufficiently multi-extremal benchmark functions.
% We would like to encourage you to list your keywords within
% the abstract section using the \keywords{...} command.
\keywords{deterministic global optimization $\cdot$ stochastic global optimization
  $\cdot$ parallel numerical methods $\cdot$ derivative-free algorithms $\cdot$ black-box
optimization}
\end{abstract}
%
\section{Introduction}

The nonlinear global optimization of the non-convex functions is considered to be one of the most
difficult problems of mathematical programming traditionally. The finding of the global minimum of
a function of several variables appears to be more difficult than the local optimization in the
thousand-dimensional space often. For the latter, the application of the simplest gradient descent
method may appear to be enough whereas in order to \textit{guarantee} the finding of the global
optimum by the optimization methods, one has to accumulate the information on the behavior of the
objective function within the whole search domain
\cite{Jones2009,Paulavicius2011,Evtushenko2013,strSergGO}. To date, various stochastic global
optimization algorithms have attracted much attention, first of all, the evolution ones
\cite{Storn1997, SCHLUTER2009, KennedyEberhart1995}. These ones have rather simple structure
and allow solving the problems of large dimensionality. However, the evolution algorithms ensure
the global convergence in the probabilistic sense only.

In the present paper, the open source implementations of nine different global optimization methods
were considered. All algorithms were tested using a set of 900 essentially multiextremal functions,
which has been generated using special problem generators \cite{Gaviano2003, grishaginClass}.
Besides the comparison of the sequential algorithms, a comparison of the MIDACO solver
\cite{Schlueter2012} with the Globalizer software system \cite{globalizerSystem,Strongin2018} has been
performed while running on Lobachevsky supercomputer using a subset of 200 test problems.

@@

Earlier, the comparison of the stochastic global optimization algorithms \cite{Ali2005, JSSv060i06}
as well as of the deterministic ones \cite{posik2012, KVASOV2018245, Liberti2005} between each
other has been considered in the literature. In these works, the majority of modern methods have
been studied in details, but the emphasis was made mainly on the sequential algorithms while the
reliability and the convergence speed were the main criteria of efficiency of the optimization
methods. In the majority of works, the sets of well-known test problems (for example, the Rastrigin
functions, Ackley function, etc.) were taken as the sets of test functions. The sizes of such sets don't
exceed 100 different functions usually, some of which can be the single-extremal ones (such as the
Rosenbrock functions).

@@

Using the multiple mapping allows solving initial problem (\ref{eq:task}) by parallel solving the
problems
\[
\min\{\varphi(y^s(x)):x\in [0,1]\}, 1\leqslant s\leqslant S
\]
on a set of intervals $[0,1]$ by the index method. Each one-dimensional problem is solved on a
separate processor. The trial results at the point \(x^k\) obtained for the problem being solved by
particular processor are interpreted as the results of the trials in the rest problems (in the
corresponding points \(x^{k_1},\dots,x^{k_S})\). In this approach, a trial at the point \(x^k \in
[0,1]\) executed in the framework of the \(s^{th}\) problem, consists in the following sequence
of
operations:
\par
1. Determine the image \(y^k=y^s (x^k)\) for the evolvent \(y^s (x)\).
\par
2. Inform the rest of processors about the start of the trial execution at the point \( y^k\) (the
blocking of the point \(y^k\) ).
\par
3. Determine the preimages \(x{}^{k_s}  \in [0,1], 1\leqslant s\leqslant S\), of the point \(y^k\) and
interpret the
trial executed at the point \(y^k \in D \) as the execution of the trials in the \(S\) points
\(x{}^{k_1} ,\dots,x{}^{k_s} \)
\par
4. Inform the rest of processors about the trial results at the point \(y^k\).
\par
The decision rules for the mentioned parallel algorithm, in general, are the same as the rules of the
sequential algorithm (except the method of the trial execution). Each processor has its own copy
of the software realizing the computations of the problem functions and the decision rule of the
index algorithm. For the organization of the interactions among the processors, the queues are
created on each processor, where the processors store the information on the executed iterations
in the form of the tuples: the processor number \(s\), the trial point \(x{}^{k_s}\).
The modifications of the method taking into account multiply criteria are given in \cite{GERGEL2017,Gergel2018}.
\par
The mentioned parallelization scheme was proposed in \cite{Gergel2009} and implemented in the
Globalizer system with the use of MPI technology. Main
features of implementation consist in the following:
\begin{itemize}
  \item A separate MPI-process is created for each
 of \(S\) one-dimensional problems being solved, usually, one process per one processor
 employed.
 \item Each process can use $p$ threads to parallel execute $p$ trials, usually one thread per an
accessible core.
\end{itemize}

@@

Let us suppose a test problem to be solved if the optimization method executes the scheduled trial
\(y^k\) in a \(\delta\)-nearness of the global minimum \(y^*\) i. e. $\left\|y^k-y^*\right\|\leqslant \delta
= 0.01\left\|b-a\right\|$ where \(a\) and \(b\) are the left and the right boundaries of the hypercube
from (\ref{eq:task}). If this relation is not fulfilled before the expiration of the limit of the number of
trials, the problem was considered to be unsolved. The maximum limit of the number of trials was set
for each problem class according to the dimensionality and complexity (see Table \ref{tab:limits}).

@@

\begin{table}
\begin{center}
\caption{Class-specific parameters of optimization algorithms}
  \begin{tabular}{|l|{c}|{c}|{c}|}
    \hline
    & AGS, AGS\(l\) & CRS & DE\\
  \hline
  \(F_{GR}\) & \(r=3\) & popsize=150 & mutation=(1.1,1.9), popsize=60 \\
  \hline
  GKLS 2d Simple & \(r=4.6\) & popsize=200 & mutation=(1.1,1.9), popsize=60 \\
  \hline
  GKLS 2d Hard & \(r=6.5\) & popsize=400 & mutation=(1.1,1.9), popsize=60 \\
  \hline
  GKLS 3d Simple & \(r=3.7\) & popsize=1000 & mutation=(1.1,1.9), popsize=70 \\
  \hline
  GKLS 3d Hard & \(r=4.4\) & popsize=2000 & mutation=(1.1,1.9), popsize=80 \\
  \hline
  GKLS 4d Simple & \(r=4.7\) & popsize=8000 & mutation=(1.1,1.9), popsize=90 \\
  \hline
  GKLS 4d Hard & \(r=4.9\) & popsize=16000 & mutation=(1.1,1.9), popsize=100 \\
  \hline
  GKLS 5d Simple & \(r=4\) & popsize=25000 & mutation=(1.1,1.9), popsize=120 \\
  \hline
  GKLS 5d Hard & \(r=4\) & popsize=30000 & mutation=(1.1,1.9), popsize=140 \\
  \hline
\end{tabular}
  \label{tab:params}
\end{center}
\end{table}

@@

Computational experiments with parallel algorithms have been carried out on the Lobachevsky
supercomputer at
State University of Nizhni Novgorod. A computational node includes 2 Intel
Sandy Bridge E5-2660 2.2 GHz processors, 64 GB RAM. Each CPU has 8 cores (i. e. total 16
cores were available per a node). All the computational nodes are working under CentOS Linux 7.2.
All the considered algorithms are implemented using C++ and compiled by GCC 4.8.5.

@@

The averaged speedup in the number of iterations and the one in time (in brackets) are presented in
Table \ref{tab:speedup}. In the first row of Table \ref{tab:speedup}, the averaged number of
iterations and the averaged time of solving the problems in the sequential mode (in brackets) are
presented. MIDACO uses a smaller number of communications in the parallel computations.
Therefore, the indicators of speedup increased monotonously with increasing number of
computational nodes employed. In Globalizer, each node should transfer the results of the executed
iterations to all other nodes. Therefore, the efficiency of parallelization decreased with increasing
number of nodes and the number of active threads on each node. This is evident from Table
\ref{tab:speedup} and Picture \ref{fig:speedup_p16}: the speedup in time decreased at the transition from 4 nodes to 8 ones when
using 16 threads. In the case of using 1 thread per a node, the speedup of both optimization methods
were approximately the same (see Picture \ref{fig:speedup_p16}). However, it should be noted that Globalizer executed less number of
trials, and spends much less time for solving all problems when \(p=1\). With increasing number of
nodes and threads, MIDACO approaches Globalizer in the averaged time of problem solving (see Pictures \ref{fig:time_p1}, \ref{fig:time_p16}),
however, its reliability decreases at the same time.

>>>>

\begin{abstract}
This work considers several stochastic and deterministic derivative-free global optimization
algorithms. In the first part of the paper popular sequential open-source solvers are compared against
Globalizer solver, which is developed at the Lobachevsky State University. The Globalizer is
designed to solve problems with black-box objectives satisfying the Lipschitz condition and shows
competitive performance with other similar solvers. The comparison is done on several sets of
challenging multi-extremal benchmark functions. The second part of this work is devoted to
comparison between the Globalizer and MIDACO solvers on systems with shared and distributed
memory. MIDACO is a state-of-the-art global solver included to the TOMLAB optimization
environment for MATLAB. Results of the benchmark show advantages of the Globalizer on small-
dimensional, but sufficiently multi-extremal benchmark functions.
% We would like to encourage you to list your keywords within
% the abstract section using the \keywords{...} command.
\keywords{deterministic global optimization $\cdot$ stochastic global optimization
  $\cdot$ parallel numerical methods $\cdot$ derivative-free algorithms $\cdot$ black-box
optimization}
\end{abstract}
%
\section{Introduction}

Nonlinear global optimization of non-convex functions is traditionally considered to be one of the most
difficult problems of mathematical programming. The finding of the global minimum of
a function of several variables often appears to be more difficult than local optimization in the
thousand-dimensional space. In local optimization applying of the simplest gradient descent
method may appear to be enough, whereas in order to \textit{guarantee} find the global
optimum, an optimization method has to accumulate information about behavior of the
objective function within the whole search domain
\cite{Jones2009,Paulavicius2011,Evtushenko2013,strSergGO}. Nowadays, various stochastic global
optimization algorithms have attracted much attention, first of all, the evolution ones
\cite{Storn1997, SCHLUTER2009, KennedyEberhart1995}. They have rather simple structure
and allow solving the problems of large dimensionality. However, the evolution algorithms ensure
the global convergence in the probabilistic sense only.

In the present paper, the open-source implementations of nine different global optimization methods
were considered. All the algorithms were tested using a set of 900 essentially multiextremal functions,
which has been generated using generators of test problems \cite{Gaviano2003, grishaginClass}.
Besides the comparison of the sequential algorithms, a comparison of the MIDACO solver
\cite{Schlueter2012} with the Globalizer software system \cite{globalizerSystem,Strongin2018} on a subset of 200 test problems has been
performed in the Lobachevsky supercomputer environment.

@@

Earlier, the comparison of the stochastic global optimization algorithms \cite{Ali2005, JSSv060i06}
as well as of the deterministic ones \cite{posik2012, KVASOV2018245, Liberti2005} between each
other has been considered in the literature. In these works, most of modern methods have
been studied in details, but the emphasis was made mainly on the sequential algorithms while the
reliability and the convergence speed were the main criteria of efficiency of the optimization
methods. In the majority of works, the sets of well-known test problems (for example, the Rastrigin
function, Ackley function, etc.) were taken as the sets of test functions. The sizes of such sets don't
exceed 100 different functions usually, some of which can be the single-extremal ones (such as the
Rosenbrock function).

@@

Using the multiple mapping allows solving initial problem (\ref{eq:task}) by parallel solving the
problems
\[
\min\{\varphi(y^s(x)):x\in [0,1]\}, 1\leq s\leq S
\]
on a set of intervals $[0,1]$ by the index method. Each one-dimensional problem is solved on a
separate processor. The trial results at the point \(x^k\) obtained for the problem being solved by
particular processor are interpreted as the results of the trials in the rest problems (in the
corresponding points \(x^{k_1},\dots,x^{k_S})\). In this approach, a trial at the point \(x^k \in
[0,1]\) executed in the framework of the \(s^{th}\) problem, consists in the following sequence
of
operations:
\par
1. Determine the image \(y^k=y^s (x^k)\) for the evolvent \(y^s (x)\).
\par
2. Inform the rest of processors about the start of the trial execution at the point \( y^k\) (the
blocking of the point \(y^k\) ).
\par
3. Determine the preimages \(x{}^{k_s}  \in [0,1], 1\leq s\leq S\), of the point \(y^k\) and
interpret the
trial executed at the point \(y^k \in D \) as the execution of the trials in the \(S\) points
\(x{}^{k_1} ,\dots,x{}^{k_s} \)
\par
4. Inform the rest of processors about the trial results at the point \(y^k\).
\par
The decision rules for the mentioned parallel algorithm, in general, are the same as the rules of the
sequential algorithm (except the method of the trial execution). Each processor has its own copy
of the software realizing the computations of the problem functions and the decision rule of the
index algorithm. For the organization of the interactions among the processors, the queues are
created on each processor, where the processors store the information on the executed iterations
in the form of the tuples: the processor number \(s\), the trial point \(x{}^{k_s}\).
The modifications of the method taking into account multiply criteria are given in \cite{GERGEL2017,Gergel2018}.
\par
The mentioned parallelization scheme was proposed in \cite{Gergel2009} and implemented in the
Globalizer system with the use of MPI technology. Main
features of implementation consist in the following:
\begin{itemize}
  \item A separate MPI-process is created for each
 of \(S\) one-dimensional problems being solved, usually, one process per one processor
 employed.
 \item Each process can use $p$ threads to parallel execute $p$ trials, usually one thread per an
accessible core.
\end{itemize}

@@

Let us suppose a test problem to be solved if the optimization method executes the scheduled trial
\(y^k\) in a \(\delta\)-nearness of the global minimum \(y^*\) i. e. $\left\|y^k-y^*\right\|\leq \delta
= 0.01\left\|b-a\right\|$ where \(a\) and \(b\) are the left and the right boundaries of the hypercube
from (\ref{eq:task}). If this relation is not fulfilled before the expiration of the limit of the number of
trials, the problem was considered to be unsolved. The maximum limit of the number of trials was set
for each problem class according to the dimensionality and complexity (see Table \ref{tab:limits}).

@@

\begin{table}
\begin{center}
\caption{Class-specific parameters of the optimization algorithms}
  \begin{tabular}{|l|{c}|{c}|{c}|}
    \hline
    & AGS, AGS\(l\) & CRS & DE\\
  \hline
  \(F_{GR}\) & \(r=3\) & popsize=150 & mutation=(1.1,1.9), popsize=60 \\
  \hline
  GKLS 2d Simple & \(r=4.6\) & popsize=200 & mutation=(1.1,1.9), popsize=60 \\
  \hline
  GKLS 2d Hard & \(r=6.5\) & popsize=400 & mutation=(1.1,1.9), popsize=60 \\
  \hline
  GKLS 3d Simple & \(r=3.7\) & popsize=1000 & mutation=(1.1,1.9), popsize=70 \\
  \hline
  GKLS 3d Hard & \(r=4.4\) & popsize=2000 & mutation=(1.1,1.9), popsize=80 \\
  \hline
  GKLS 4d Simple & \(r=4.7\) & popsize=8000 & mutation=(1.1,1.9), popsize=90 \\
  \hline
  GKLS 4d Hard & \(r=4.9\) & popsize=16000 & mutation=(1.1,1.9), popsize=100 \\
  \hline
  GKLS 5d Simple & \(r=4\) & popsize=25000 & mutation=(1.1,1.9), popsize=120 \\
  \hline
  GKLS 5d Hard & \(r=4\) & popsize=30000 & mutation=(1.1,1.9), popsize=140 \\
  \hline
\end{tabular}
  \label{tab:params}
\end{center}
\end{table}

@@

Computational experiments with parallel algorithms have been carried out on the Lobachevsky
supercomputer at State University of Nizhni Novgorod. A computational node includes 2 Intel
Sandy Bridge E5-2660 2.2 GHz processors, 64 GB RAM. Each CPU has 8 cores (i. e. total 16
cores were available per a node). All the computational nodes are working under CentOS Linux 7.2.
All the considered algorithms are implemented using C++ and compiled by the GCC 4.8.5.

@@

The averaged speedup in the number of iterations and the one in time (in brackets) are presented in
Table \ref{tab:speedup}. In the first row of Table \ref{tab:speedup}, the averaged number of
iterations and the averaged time of solving the problems in the sequential mode (in brackets) are
presented. MIDACO uses a smaller number of communications in the parallel computations than the Globalizer.
Therefore, the indicators of speedup increased monotonously with increasing number of
computational nodes employed. In Globalizer, each node should transfer the results of the executed
iterations to all other nodes. Therefore, the efficiency of parallelization decreased with increasing
number of nodes and the number of active threads on each node. This is evident from Table
\ref{tab:speedup} and Picture \ref{fig:speedup_p16}: the speedup in time decreased at the transition from 4 nodes to 8 ones when
using 16 threads. In the case of using 1 thread per a node, the speedups of both optimization methods
were approximately the same (see Picture \ref{fig:speedup_p16}). However, it should be noted that Globalizer executed less number of
trials, and spends much less time for solving all problems when \(p=1\). With increasing number of
nodes and threads, MIDACO approaches Globalizer in the averaged time of problem solving (see Pictures \ref{fig:time_p1}, \ref{fig:time_p16}),
however, its reliability decreases at the same time.
